---
title: Record Linkage - generate data for Fellegi-Sunter Model
author: Zygmunt Zawadzki
date: '2020-10-01'
slug: record-linkage-generate-data-for-fellegi-sunter-model
tags:
  - record-linkage
  - rpkg-reclin
---

I'm a little fascinated with the record linkage topic. There are some nice ideas in that area. For example - the Fellegi-Sunter Model which uses the M and U probabilities to calculate a score which can make the matching. And the best part is the EM algorithm to estimate M and U with no external knowledge (you need not know which pairs match, no training data, etc).

<!--more-->

When the assumptions are met, it's obvious that the method must work. However, to better understand some method I like to simulate the data that follows specific model. Then I can make simulations and play with assumptions (e.g. what if I relax the assumption of independence, etc). After some experiments, I have a much better understanding in a topic.

This time I played with Fellegi-Sunter.

```{r}
library(dplyr)
library(reclin)
library(customLayout)

set.seed(123)
N <- 1000

## nn - number of different values for given column
## 1/nn is the U probability
nn <- c(2,10,100,500, 1000)
mProbs <- c(0.99, 0.99, 0.97, 0.95, 0.90)

simulation <- function(N, nn, mProbs) {

  tbl1 <- lapply(nn, function(n)
         sample.int(n, size = N, replace = TRUE)
  ) %>% as.data.frame() %>% as_tibble()
  colnames(tbl1) <- paste0("v", seq_along(tbl1))

  # copy the data
  tbl2 <- tbl1

  # adjust each column in a loop using m-probability
  for(i in seq_along(tbl1)) {
    
    # replaces value with different value when sampled value above M prob
    p <- runif(N) > mProbs[[i]] 
    while (any(p)) {
      # make sure that different value will be selected
      newVals <- sample.int(nn[i], sum(p), replace = TRUE)
      org <- tbl2[[i]][p]
      tbl2[[i]][p] <- newVals
      p[p][org != newVals] <- FALSE
    }
  }

  pairs <- pair_blocking(tbl1, tbl2)
  pairs <- compare_pairs(pairs, colnames(tbl1))
  model <- problink_em(pairs)
  model
}

result <- simulation(N, nn, mProbs)
result
```

### Distribution of estimates:

```{r}


simAll <- replicate(50, simulation(N, nn, mProbs), simplify = FALSE)

mAll <- t(sapply(simAll, "[[", "mprobs"))

plot_hist <- function(i, p, title = "M-prob") {
  hist(unlist(mAll[,i]), main = paste0("V", i), xlab = title)
  abline(v = p, lwd = 4, col = "red", lty = 2)
  invisible()
}


lay <- customLayout::lay_bind_row(
  customLayout::lay_new(cbind(1,2,3)),
  customLayout::lay_new(cbind(1,2))
)

lay_set(lay)
invisible(mapply(plot_hist, seq_along(mProbs), mProbs))
```

# Bibliography

If you want to know more about record linkage, there're some good resources:

1. Loo M PJ ,van,der. The stringdist Package for Approximate String Matching. The R Journal [Internet]. 2014 [cited 2020 Apr 13];6(1):111. Available from: https://journal.r-project.org/archive/2014/RJ-2014-011/index.html
2. Sariyar M, Borg A. The RecordLinkage Package: Detecting Errors in Data. The R Journal [Internet]. 2010 [cited 2020 Apr 13];2(2):61–7. Available from: https://journal.r-project.org/archive/2010/RJ-2010-017/index.html
3. Micro-Fusion-07-M-Fellegi-Sunter and Jaro Approach v1.0_0.pdf [Internet]. [cited 2020 Sep 25]. Available from: https://ec.europa.eu/eurostat/cros/system/files/Micro-Fusion-07-M-Fellegi-Sunter%20and%20Jaro%20Approach%20v1.0_0.pdf
4. Harron K, Goldstein H, Dibben C. Methodological developments in data linkage. Chichester, West Sussex, United Kingdom: Wiley; 2016. 
5. Herzog TN, Scheuren FJ, Winkler WE. Data quality and record linkage techniques. New York, NY: Springer; 2007. 227 p. 
6. Christen P. Data matching: concepts and techniques for record linkage, entity resolution, and duplicate detection. Berlin ; New York: Springer; 2012. 270 p. (Data-centric systems and applications). 
7. McDonald M. An Introduction to Probabilistic Record Linkage. :43. 
