---
title: Make asynchronous call in shiny.
author: Zygmunt Zawadzki
date: '2017-04-24'
slug: ''
categories: 
  - R
  - shiny
tags:
  - R
  - shiny
  - future
---

Sometimes when you are creating shiny's application, and you need to allow the user to perform a long computation (e.g. some model fitting). However, time-demanding calculations are quite problematic, because the whole session needs to wait until it finishes, blocking any other work in different part of the system. 

Fortunately, there is a way to perform an asynchronous job in shiny, so the long-running task can be resolved in the background without blocking the application. To take advantage of that feature we need to learn something about the future package.

The future allows to schedule some code to evaluation in other R's thread in a very straightforward way. To detailed explanation, please read this vignette https://cran.r-project.org/web/packages/future/vignettes/future-1-overview.html. In this article, we need just some most basic functionalities.

Firstly we need to load the package and create an execution plan. See the following code:

```{r}
library(future)
plan(multiprocess)

a <- 10

# Future automatically exports all needed variables.
ft <- future({
  Sys.sleep(1)
  a * 10
})

# Checks if the value is ready.
resolved(ft)

# Overwrite the value of "a".
a <- 0
print(paste("a is equal:", a))
# Future's result is based on the old value of the "a" variable:
print(paste("future's value is equal:", value(ft)))
```

Ok. We know how to run an asynchronous job in regular R session. However, we cannot use that strategy inside shiny, because we still need to wait inside reactive to return the value.  Fortunately shiny has an `invalidateLater` functionality, which allows us to schedule current reactive context to re-execution. 

So we can use a very simple pattern - reactiveValues is used to store currently executed future's call.  If it is value is equal to null, that mean there's no running job so that we can schedule a new one. If the value is still unresolved, we use *invalidateLater* to schedule re-execution after few moments. When the job is ready, we can return the value. See the following listing for the simple example:

```{r, eval=FALSE}
future <- reactiveValues(future = NULL)

output$FuturePlot <- renderPlot({
   input$MakePlot
   
   x <- NULL
   
  isolate({
    
    if(is.null(future$future)) {
      future$future <- future({
        Sys.sleep(5)
        cbind(rnorm(1000), rnorm(1000))
      })
      cat("Work scheduled\n")
    } else if(!resolved(future$future)) {
      cat("Waiting\n")
    } else {
      x <- value(future$future)
      future$future <- NULL
      cat("Done!\n")
    }
  })
  
  if(is.null(x)) invalidateLater(1000) # Wait 1s
  
  req(x) # Stop there if x is still null
  plot(x)
})

```

## Putting it all together.

There's the working app:

```{r, eval=FALSE}
# Define UI for application that draws a histogram
ui <- fluidPage(
   
   # Application title
   titlePanel("Old Faithful Geyser Data"),
   
   # Sidebar with a slider input for number of bins 
   sidebarLayout(
      sidebarPanel(
         sliderInput("bins",
                     "Number of bins:",
                     min = 1,
                     max = 50,
                     value = 30),
        actionButton("MakePlot", "MakePlot")
      ),
      
      # Show a plot of the generated distribution
      mainPanel(
         fluidRow(
           column(6, plotOutput("distPlot")),
           column(6, plotOutput("FuturePlot")))
      )
   )
)

# Define server logic required to draw a histogram
server <- function(input, output) {
   
   output$distPlot <- renderPlot({
      # generate bins based on input$bins from ui.R
      x    <- faithful[, 2] 
      bins <- seq(min(x), max(x), length.out = input$bins + 1)
      
      # draw the histogram with the specified number of bins
      hist(x, breaks = bins, col = 'darkgray', border = 'white')
   })
   
   
   future <- reactiveValues(future = NULL)
   
   output$FuturePlot <- renderPlot({
     input$MakePlot
     
     x <- NULL
     
    isolate({
      
      if(is.null(future$future)) {
        future$future <- future({
          Sys.sleep(5)
          cbind(rnorm(1000), rnorm(1000))
        })
        cat("Work scheduled\n")
      } else if(!resolved(future$future)) {
        cat("Waiting\n")
      } else {
        x <- value(future$future)
        future$future <- NULL
        cat("Done!\n")
      }
    })
    
    if(is.null(x)) invalidateLater(1000) # Wait 1s
    
    req(x) # Stop there if x is still null
    plot(x)
  })
   
}

shinyApp(ui = ui, server = server)
```

I hope you enjoyed that post, and found it usefull. Welcome in my blog:)
