---
title: "Updated vectors in R 3.4.0"
date: '2017-05-06'
author: Zygmunt Zawadzki
output: html_document
categories: 
  - R
tags:
  - performance
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "CairoSVG", echo = FALSE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4)
```

The last version of R comes with a small update of vector behavior in regard to inserting new value just after the last element. See the entry from the announcement (https://stat.ethz.ch/pipermail/r-announce/2017/000612.html):

```
* Assigning to an element of a vector beyond the current length now
     over-allocates by a small fraction. The new vector is marked
     internally as growable, and the true length of the new vector is
     stored in the truelength field. This makes building up a vector
     result by assigning to the next element beyond the current length
     more efficient, though pre-allocating is still preferred.  The
     implementation is subject to change and not intended to be used
     in packages at this time.
```

In this post, I'll try to show how does this work in practice, and what does it mean to the user.

## Functions used in benchmark

To test the new functionality I created three simple functions which add a new element at the end of the vector in three different ways:

- Assign new item using `[[]]` operator:

```{r, echo=TRUE}
fnc_assign <- function(y) {
  x <- NULL
  for (i in y) {
    x[[i]] <- i
  }
  x
}
```

- Use `c(...)` function:

```{r, echo=TRUE}
fnc_combines <- function(y) {
  x <- NULL
  for (i in y) {
    x <- c(x, i)
  }
  x
}
```

- Pre-allocate memory (recommended way):

```{r, echo=TRUE}
fnc_prealloc <- function(y) {
  x <- numeric(length(y))
  for (i in y) {
    x[[i]] <- i
  }
  x
}
```


Then I created simple bash scripts to download and compile two versions of R. The testing functions were executed in both releases using `microbenchmark` (the code used to create this analysis is available in GitHub repository - https://github.com/zzawadz/blog-entry-vector-speed).

## Results

```{r}
library(ggplot2)
library(dplyr)
library(Cairo)
library(magrittr)
library(knitr)

files <- dir("../../data/data-2017-05-06/", full.names = TRUE)

size.idx <- grep("size", files)

files.time <- files[-size.idx]
files.size <- files[size.idx]

read_data <- function(files) {
  result <- lapply(files, function(x) {
    name <- tail(strsplit(basename(x), split = "-")[[1]], 1)
    name <- gsub(name, pattern = ".txt", replacement = "")
    name <- paste0("R-", name)
    tbl <- read.table(x)
    tbl %<>% mutate(Time = Time / 1e6)
    cbind(Version = name, tbl)
  })

  Reduce(rbind, result)
}
```

```{r}
time.table <- read_data(files.time)
# time.table <-
#   time.table %>% group_by(Fnc, Version) %>% filter(abs(Time - median(Time)) < 1.5 * IQR(Time))
  
make_plot <- function(dt){
  p <- ggplot(dt) + geom_boxplot(aes(fill = Version, y = Time, x = Fnc))
  p + scale_fill_manual(values = c("R-3.3.3" = "#E41A1C", "R-3.4.0" = "#377EB8"))
}

make_table <- function(dt) {
  dt <- dt %>% group_by(Version, Fnc)  %>% summarise("Median Time" = median(Time))
  dt <- dt %>% ungroup %>% mutate(Ratio = `Median Time` / min(`Median Time`))
  kable(dt, digits = 2)
}


```

### Assign with `[[]]` operator.

In the first version of the append function, the difference is significant. For $10^4$ elements, the speed up is by order of magnitude. 

```{r}
make_plot(time.table %>% filter(Fnc == "fnc_assign(y)"))
make_table(time.table %>% filter(Fnc == "fnc_assign(y)"))
```

Even when we compare this result with the pre-allocated version, it is only slightly slower. Good news!

```{r}
tbl <- time.table %>% filter(Fnc %in% c("fnc_assign(y)", "fnc_prealloc(y)"), Version == "R-3.4.0")

make_plot(tbl) + ylim(c(0,10))
make_table(tbl)
```



### Pre-allocation.

No real difference;)

```{r}
make_plot(time.table %>% filter(Fnc == "fnc_prealloc(y)")) + ylim(c(5,12))
make_table(time.table %>% filter(Fnc == "fnc_prealloc(y)"))
```

### c() function.

This version does not support the new functionality. It might be a bit confusing because it seems to be quite natural to add the new observation to the end of a vector using `c(...)` function. However, this causes massive loss of performance, and it is a big mistake! **It's better to avoid** `c(...)` **in the** `for` **loop!**

```{r}
make_plot(time.table %>% filter(Fnc == "fnc_combines(y)"))
make_table(time.table %>% filter(Fnc == "fnc_combines(y)"))
```


## Impact of vector size.

It is worth to note that in the old version (and in the current one when the `c` function is used) creating a vector of length $n$ by adding just one element in the loop has $O(n^2)$ complexity. Inserting just one element causes reallocating the whole vector in the memory, so in the ith iteration, we need to move $i-1$ elements. So in the end we get $1 + 2 + 3 + ... + n - 2 + n - 1$ operations, which is $O(n^2)$.

In the new version, the reallocation can be done less often so it can be much more linear. See the following graphs for comparison:

```{r}
size.table <- read_data(files.size)

size.time <- size.table %>% group_by(Version, Fnc) %>% summarise(Mean = mean(Time))

size.time$Size <- as.character(size.time$Fnc) %>% stringi::stri_extract_all_regex("[0-9]+", simplify = TRUE) %>% as.numeric()

size.plot <- ggplot(size.time) + geom_point(aes(x = Size, y = Mean, color = Version), size = 3) + xlab("Vector size") + ylab("Time in miliseconds") + theme_bw(16) + scale_color_manual(values = c("R-3.3.3" = "#E41A1C", "R-3.4.0" = "#377EB8"))
size.plot
```

And the zoom for the new version:

```{r}
r34.size <- size.time %>% filter(Version == "R-3.4.0")
r34.size.plot <- ggplot(r34.size) + geom_point(aes(x = Size, y = Mean, color = Version), size = 3) + xlab("Vector size") + ylab("Time in miliseconds") + theme_bw(16)+ scale_color_manual(values = c("R-3.3.3" = "#E41A1C", "R-3.4.0" = "#377EB8"))

r34.size.plot
```

## Summary

In R 3.4.0 the memory management for vectors is much better organized so resizing of the vector is much faster ($O(n)$ vs. $O(n^2)$). The only problem is that this does not apply to the `x <- c(x, i)` - this one is always slow. But the best option is to stay with preallocation because it is still the fastest way of dealing with growable vectors:)
