---
title: DepthProc hit 20k downloads.
date: '2018-06-11'
tags:
  - R
  - R Packages
  - rpkg-depthproc
---

My first package published on CRAN - DepthProc recently hit 20k downloads.

```{r, cache=TRUE}
library(cranlogs)
library(ggplot2)
downloads <-
cran_downloads("DepthProc", from = "2014-08-21", to = "2018-06-10")

ggplot(downloads) +
  geom_line(aes(x = date, y = cumsum(count))) +
  ylab("Downloads") + xlab("Date") + theme_bw() +
  ggtitle("DepthProc", "Download stats")

```

There are some jumps on the line. I wondered if they all occurred just after the package release (old users updates to the new versions). Here's some code to check this.

```{r, message=FALSE, warning=FALSE, cache=TRUE}
library(rvest)
library(lubridate)
library(dplyr)
path <- "https://cran.r-project.org/src/contrib/Archive/DepthProc/"

versionDate <- html_table(read_html(path))[[1]]
versionDate <- versionDate[,c(2,3)]
versionDate <- versionDate %>% 
  filter(grepl(Name, pattern = "DepthProc")) %>% 
  mutate(Date = ymd(substring(`Last modified`, 1, 10))) %>% 
  select(Date)

versionDate <- bind_rows(versionDate, 
        data_frame(Date = ymd("2018-06-05"))) # Add current release
versionDate

ggplot() + 
  geom_line(data = downloads, aes(x = date, y = cumsum(count))) + 
  ylab("Downloads") + xlab("Date") + theme_bw() + 
  ggtitle("DepthProc", "Download stats (vertical bars indicates new versions)") +
  geom_vline(data = versionDate, 
      aes(xintercept = Date), color = "darkgreen", linetype = "dashed")

```

It seems that the main jump was related to something else. Anyway, it's good to see that people use your work:)

If you have any questions about the package please send me an email or post an issue on the package's GitHub page (https://github.com/zzawadz/DepthProc/issues).

### Useful links:

- https://www.depthproc.zstat.pl/ - the package website (built with `pkgdown`).
- https://arxiv.org/abs/1408.4542 - description of the package's capabilities.
