---
title: X-ray your data
author: Zygmunt Zawadzki
date: '2018-06-23'
slug: x-ray-your-data
tags:
  - R
  - Data cleansing
---

When I was checking my notes, I found a piece of information about `xray` package. It is pretty simple, only exports three functions, but all of them are quite useful.

Search for the most common problems.
The first function `anomalies`, reports some statistics regarding the most basic data problems:

- NA.
- Zeroes.
- A categorical variable with only one level.
- Blank strings.


I usually test the data for the presence of NAs using some simple code like
`sapply(data, function(x) mean(is.na(x)))`, but I'll happily switch to xray, because with one line it performs more tests, and the output is far more nicely formatted.

See the example below:

```{r}
# install.packages("xray")
library(xray)

anom <- anomalies(iris)
anom

iris2 <- iris

iris2$Petal.Length[sample.int(150, 80)] <- 0
iris2$Sepal.Width[sample.int(150, 80)] <- ifelse(rbinom(80, size = 1, prob = 0.3) == 0, Inf, -Inf)
iris2$SpeciesChar <- ifelse(rbinom(150, size = 1, prob = 0.9) == 0, as.character(iris2$Species), "")
iris2$Species[sample.int(150, 20)] <- NA

anom2 <- xray::anomalies(iris2)
anom2
```

The second function `distributions` plots distributions for each variable, and returns a summary table. It's also a quite common step, so I'm glad that it's automated in the package, and the output is also beautifully formatted.

The next example:

```{r dist}
distributions(iris2) # as simple as this
```

The last function is used to check the data over time. The example usage can be found in the https://github.com/sicarul/xray/, so check the package's website.

### Summary.

This package inspired me to the further research, for other tools to validate and prepare your data to the actual analysis, and I found a real pearl - the `vtreat` package. It seems to be a real swiss-army knife for preparing data for predictive modeling. But it's a topic for another post;)

### Links:
- https://blog.datascienceheroes.com/x-ray-vision-on-your-datasets/ - short description of the package.
- https://github.com/sicarul/xray/ - xray's GitHub webpage.
- https://winvector.github.io/vtreat/ - swiss-army knife for preparing your data for modeling.
